---
title: "Cultural and Cognitive Sociology"
subtitle: "A Graduate Seminar with Methods Laboratory"
author: "MA Graduate Seminar"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [references.bib]
biblio-style: apalike
link-citations: yes
description: |
  A rigorous, methodologically integrated graduate course in Cultural and
  Cognitive Sociology. Covers theory, quantitative methods (LCA, EFA/CFA,
  SEM, MCA, regression), qualitative methods, and computational approaches.
  Built for MA students with thesis and doctoral development in mind.
url: ''
cover-image: ''
github-repo: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 8,
  fig.height = 5,
  fig.align  = "center",
  cache      = FALSE,
  tidy       = FALSE
)

# Reproducibility
set.seed(20240101)
```

# Preface {-}

## How This Course Is Organized {-}

This course runs on three interlocking tracks across fifteen weeks:

**Track A — Theory.** The conceptual architecture of cultural and cognitive
sociology: what the field claims, what it disputes, and where its active
frontiers are.

**Track B — Methods Laboratory.** Operational implementation of the methods
used to study cultural and cognitive phenomena. Every theoretical module has
a corresponding lab. Code is written to run, not to illustrate.

**Track C — Research Development.** Four cumulative assignments scaffold a
thesis-quality research design across the semester. The final deliverable
(Assignment 4) is a complete research design paper, 7,000–10,000 words.

The course does not treat theory and methods as separable. Method selection
is a theoretical act. Operationalization decisions carry ontological
commitments. These connections are made explicit throughout.

---

## Pedagogical Principles {-}

1. **Constructs before measures.** Every quantitative exercise begins with
   the theoretical construct, then asks how (and whether) it can be
   operationalized with the available data.

2. **Replication before innovation.** Lab exercises begin by replicating
   published results. Only once a method is understood as implemented do
   students modify and extend it.

3. **Failure is data.** When a model fits poorly, when a factor structure
   does not replicate, when a class solution is theoretically uninterpretable
   — this is informative, not a mistake to be hidden.

4. **Pre-registration as discipline.** Assignment 3 requires a pre-registration
   document before analysis. This trains the habit of separating confirmatory
   from exploratory inference.

---

## Prerequisites {-}

Students should arrive with:

- Introductory graduate statistics (regression, basic probability)
- Working R installation (≥ 4.2.0)
- Familiarity with `tidyverse` data manipulation

No prior SEM, LCA, or MCA is assumed. These are built from scratch in the
methods labs.

Run the following to install all packages used in this course:

```{r install-packages, eval=FALSE}
source("packages.R")
```

---

## Datasets Used {-}

| Dataset | Source | Access |
|---------|--------|--------|
| World Values Survey (WVS) Wave 7 | worldvaluessurvey.org | Free registration |
| European Social Survey (ESS) Round 10 | europeansocialsurvey.org | Free registration |
| General Social Survey (GSS) 2022 | gss.norc.org | Free download |
| ISSP Religion Module III | issp.org | Free download |
| Google Books Ngrams | books.google.com/ngrams | API / bulk download |
| `poLCA` built-in datasets | R package | Installed with package |
| `lavaan` built-in datasets | R package | Installed with package |

Where external data require registration, the lab sections provide
code to load and clean the data once downloaded, plus synthetic
equivalents that mirror the variable structure for in-class use.

---

## Assignment Overview {-}

| Assignment | Due | Weight | Description |
|------------|-----|--------|-------------|
| A1: Conceptual Map | Week 3 | 15% | Trace one construct across five empirical studies |
| A2: Data Exploration | Week 6 | 20% | Research question + preliminary descriptives from a real dataset |
| A3: Pre-Registration | Week 10 | 25% | Full analytic plan, pre-specified, OSF-format |
| A4: Research Design Paper | Week 15 | 40% | 7,000–10,000 words; complete design or partial empirical analysis |

---

## A Note on Reproducibility {-}

All lab code sets a random seed explicitly. Results should be fully
reproducible across machines given the same package versions. Session
information is printed at the end of each chapter. If your results
diverge, check package versions first.

```{r session-info-preface, collapse=TRUE}
R.version.string
```
